The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts

### 2020-05-08 write-up/covid-writeup.tex
Update covid-writeup.tex

 \documentclass[11pt]{amsart}

@@

\usepackage{biblatex}
%\bibliographystyle{apalike}

@@

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{soul,color,hyperref}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage[margin=1.5in]{geometry}

@@

\begin{document}
\title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts}
% \author{Walter Dempsey\thanks{Department of Biostatistics, University of Michigan, 1415 Washington Heights, U.S.A. E-mail: wdem@umich.edu}}
\maketitle

\begin{abstract}
Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts. Limited testing capacity across many states has been widely identified as a key driver of suppressed coronavirus case-counts.  Calls for increased testing capacity are well-justified and increasingly frequent.  While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.  Moreover, tests are imperfect.  False positive/negative rates interact in complex ways with selection bias.  This note attempts to clarify this interaction.  Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error. The discussion guides a series of suggestions to improve current case-count reporting.
\end{abstract}

@@

Despite these necessary steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by Johns Hopkins University (\url{https://bit.ly/2UqFSuA}) and the New York Times (\url{https://bit.ly/2vUHfrK}) provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists the world over have been using this granular data to build model-based forecasts of future COVID-19 case-counts, deaths, and hospitalizations.  Variations on the SIR models have been used to draw inferences and build forecasts.  In most cases, parameter inference uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic (cites).

The goal of this paper is to express reservations at the use of case-counts as a proxy for prevalence and disease trajectory as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.  The reason is simple: current models ignore selection bias and measurement-error.

Selection bias enters due to differences in testing strategies across countries and states.  In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only  high-risk populations.  Testing requires the individual to self-select into testing.  Demands for testing increases, while laudable, ignore the issues of self-selection and measurement-error. While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.  We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a deleterious effect in our ability to estimate quantities of interest such as prevalence and effective reproduction number.

In this paper, we will demonstrate the complex interaction between these two fundamental concepts.  Measurement-error is often associated with parameter attenuation; howeve,r we show that the combination with selection effects can cause bias in parameter estimates to change sign and increase/decrease in magnitude.  Moreover, the bias may depend on the sampling fraction, prevalence, and population size.  Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.   We will show, through simple calculations, how we can misguide ourselves and the need for more precision and care in presenting COVD-19 statistical models to the broader research and non-academic communities.

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical argument, we demonstrate five important, yet often forgotten, tenants of statistical science.  First we show that unadjusted prevalence rates are unsurprisingly biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates incorporate an additional term in the error decomposition that highlights the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts are often an underestimate of the true rate. This implies use of observed rates will lead a data analyst to be overly optimistic in the early stages that the disease is not an issue, and also overly optimistic in the decline of the prevalence after the peak.   Fourth, we show that estimates of the reproducility rate are also impacted.

@@

>>>>

\documentclass[aoas]{amsart}

@@

\usepackage[foot]{amsaddr}
\usepackage{biblatex}
%\bibliographystyle{apalike}

@@

\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{setspace}
\DeclareGraphicsExtensions{.eps, .ps}
\usepackage{soul,color,hyperref}
\usepackage{amsmath, amsthm, amsfonts}
% \usepackage[margin=1.5in]{geometry}

@@

\begin{document}
\title[The Hypothesis of Testing]{The Hypothesis of Testing: Paradoxes arising out of reported coronavirus case-counts} %

\author{Walter Dempsey}
\address{Department of Biostatistics, University of Michigan, Ann Arbor, MI 48109}
% \email{wdem@umich.edu}

\begin{abstract}
Many statisticians, epidemiologists, economists and data scientists have registered serious reservations regarding the reported coronavirus case-counts. Limited testing capacity across many states has been widely identified as a key driver of suppressed coronavirus case-counts.  Calls for increased testing capacity are well-justified and increasingly frequent.  While expanded testing is a laudable goal, selection bias will impact estimates of disease prevalence and the effective reproduction number until the entire population is sampled.  Moreover, tests are imperfect.  False positive/negative rates interact in complex ways with selection bias.  In this paper, we attempt to clarify this interaction.  Through simple calculations, we demonstrate pitfalls and paradoxes that can arise when considering case-count data in the presence of selection bias and measurement-error. The discussion guides a series of suggestions to improve current case-count reporting.
\end{abstract}

\maketitle

@@

Despite these steps, the data landscape for understanding COVID-19 remains limited.  Public databases maintained by \href{https://bit.ly/2UqFSuA}{Johns Hopkins University} (cite) and the \href{https://bit.ly/2vUHfrK}{New York Times} (cite) provide incoming county-level information of confirmed cases and deaths.  Statisticians, epidemiologists, economists, and data scientists have all used this granular data to forecast COVID-19 case-counts, deaths, and hospitalizations (cite).  Variations on the SIR models (cite) have been used to draw inferences about infection rates and intensity.  Parameter inference most often uses observed case-count and/or observed COVID-19 related deaths to infer latent trajectories of the pandemic (cites).

The goal of this paper is to express reservations at the use of case-counts as a proxy for disease prevalence and intensity as well as its use as direct input into estimation of standard epidemiological models for inference and forecasting.  The reason is simple: current models do not account simultaneously for selection bias and measurement-error. Selection bias enters due to differences in testing strategies across countries and states.  In the US, for example, limited testing capacity has caused local and state health departments to focus on testing only high-risk populations.  Moreover, testing requires the individual to self-select into receiving the test.  Demands for increased testing capacity, while laudable, ignore the issues of self-selection and measurement-error. While increasing testing capacity increases data quantity, there is no guarantee for increased data quality.  We will show that aggressive pushes for ramped up testing capacity that are tied to decreases in data quality may have a very mitigated impact  in our ability to estimate quantities of interest such as prevalence and effective reproduction number.

In this paper, we will demonstrate the complex interaction between these two fundamental concepts.  Measurement-error is often associated with parameter attenuation; we show, however, that the combination with selection effects can cause bias in parameter estimates to change sign and increase/decrease in magnitude.  Moreover, the bias may depend on the sampling fraction, prevalence, and population size.  Without a complete understanding of these interations, we are destined to misinterpret case-counts and come to erroneous conclusions.   We will show, through simple calculations, why there is a strong need for more precision and care in presenting COVID-19 analyses to the broader research and non-academic communities.

@@

This article discusses the relationship between three statistical concepts: selection bias, measurement-error, and the long-forgotten population size. We clarify mathematically why the situation is much more complex than it first appears.  Through simple mathematical argument, we demonstrate five important, yet often forgotten, tenants of statistical science.  First we show that unadjusted prevalence rates are unsurprisingly biased when tests are imperfect. What is surprising is that the direction and magnitude of the bias can vary substantially and interacts with the sampling fraction.  Next we show that data quality for adjusted prevalence rates incorporate an additional term in the error decomposition that highlights the interplay among measurement error, selection bias, and prevalence.  Third, we show that daily trends in rate of positive tests is equally problematic even if testing rates and selection protocols are held constant.  We show that the rate of change in COVID-19 observed case-counts often first overestimate and then underestimate the true rate of change. This implies use of observed rates will lead a data analyst to be overly pessimistic in the early stages leading up to the peak time, and overly optimistic in its immediate aftermath.  We show that estimates of the effective reproduction number are similarly impacted.  Fourth, we show that cross-country comparisons are difficult at best with population size, sampling fraction, and data quality all interacting to impact null hypothesis testing.  Fifth, we discuss a simple sensitivity analysis tool for estimating data quality and therefore understanding the potential impact on case-count analyses.

@@

Social capital and models.
